Supplementary Material

### 2022-06-02 analysis/paper/abstract.Rmd
incorporate reviewer comments and resubmit

Dental calculus has proven to contain a wealth of information on the
dietary habits of past populations. These insights have, to a large
extent, been obtained by the extraction and identification of starch
granules contained within the mineralised dental plaque from a wide
range of regions and time periods. The scope of previous studies have
been limited to microfossil extraction and identification to reconstruct
dietary preferences from the archaeological record, and few studies have
attempted to address the biases of starch retention in dental calculus.
Those that have considered this problem have been limited to \emph{in
vivo} studies on modern humans and non-human primates. Here, we present
a multispecies oral biofilm model, which allows experimental research on
starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were
exposed to treatment solutions with known quantities of dietary starches
(wheat and potato) during the 25-day growth period. After this, the
starch granules were extracted from the mature biofilm (by dissolution
in EDTA), and counted. We show that the granule counts extracted from
the model dental calculus represented a low proportion (ranging from
0.06\% to 0.16\%) of the total number of granules exposed to the
biofilms throughout the experiment. Additionally, we found that the
ratios of granule sizes from the extracted starch granules differed from
the original treatment solutions, with large granules (\textgreater20
\(\mu\)m) consistently being under-represented. We also found a
positive correlation between the absolute granule counts and dry-weight of the
biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), meaning the absolute
quantity of starch granules will increase as the size of the calculus deposit
increases. A similar, but weaker correlation was found between
the concentration (count per mg) of granules and dry-weight (\emph{r} =
0.3, 90\%CI{[}0.0618, 0.506{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that
dental calculus presents a very small, and partly biased picture of the
original dietary intake of starches, with an over-representation of
plants producing granules smaller than 20 \(\mu\)m in size. The
experimental model presented here is well-suited to address the need for
further validation of methods and biases associated with dietary
research on dental calculus.

>>>>

Dental calculus has proven to contain a wealth of information on the
dietary habits of past populations. These insights have, to a large
extent, been obtained by the extraction and identification of starch
granules contained within the mineralised dental plaque from a wide
range of regions and time periods. The scope of previous studies have
been limited to microfossil extraction and identification to reconstruct
dietary preferences from the archaeological record, and few studies have
attempted to address the biases of starch retention in dental calculus.
Those that have considered this problem have been limited to \emph{in
vivo} studies on modern humans and non-human primates. Here, we present
a multispecies oral biofilm model, which allows experimental research on
starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were
exposed to treatment solutions with known quantities of dietary starches
(wheat and potato) during the 25-day growth period. After this, the
starch granules were extracted from the mature biofilm (by dissolution
in EDTA), and counted. We show that the granule counts extracted from
the model dental calculus represented a low proportion (ranging from
0.06\% to 0.16\%) of the total number of granules exposed to the
biofilms throughout the experiment. Additionally, we found that the
ratios of granule sizes from the extracted starch granules differed from
the original treatment solutions, with large granules (\textgreater20
\(\mu\)m) consistently being under-represented. We also found a
positive correlation between the absolute granule counts and dry-weight of the
biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), meaning the absolute
quantity of starch granules will increase as the size of the calculus deposit
increases. A similar, but weaker correlation was found between
the concentration (count per mg) of granules and dry-weight (\emph{r} =
0.3, 90\%CI{[}0.0618, 0.506{]}).\\
Our results complement and reinforce previous \emph{in vivo} studies suggesting
that dental calculus presents a very small, and partly biased picture of the
original dietary intake of starches, with an over-representation of
plants producing granules smaller than 20 \(\mu\)m in size. The
experimental model presented here is well-suited to address the need for
further validation of methods and biases associated with dietary
research on dental calculus.

### 2022-06-02 analysis/paper/conclusion.Rmd
incorporate reviewer comments and resubmit

This preliminary study shows that a very small proportion of the input starch
granules are retained in a dental calculus model. This and previous studies 
have shown that calculus has a low capacity for retention of starch granules, 
an effect that is compounded by diagenetic effects in archaeological remains, 
resulting in low overall counts of extracted granules. 
The proportion of starches consumed will in many cases be reflected 
in the quantity
of starches extracted from the dental calculus---i.e., the more starch granules
entering the oral cavity, the more will be recovered from extraction---at 
least in modern calculus samples unaffected by diagenesis and hydrolysis.
Whether or not this also applies to archaeological samples remains to be tested.
Additionally, we have
shown that the size of granules will influence the likelihood of incorporation, 
as large (>20 $\mu$m) starches have a decreased incorporation rate, medium 
(10--20 $\mu$m)
starches an increased rate, and small (<10 $\mu$m) granules remained somewhat
constant. The size of calculus deposit also seems to influence the capacity of 
granule incorporation; as the size of the deposit increases, so does the
absolute count of incorporated granules.  
While we have shown multiple factors that influence the likelihood 
of incorporation, the process still appears to be somewhat stochastic. Further
research is needed to make sense of the contributing factors, and to explore the
mechanisms of intra-oral starch incorporation and retention in dental calculus.
The oral biofilm model described in this study provides a method
to explore the incorporation and extraction of dietary compounds from dental calculus 
in a controlled laboratory setting, and unearth the potential biases associated 
with dietary research conducted on archaeological dental calculus.

>>>>

This preliminary study shows that a very small proportion of the input starch
granules are retained in a dental calculus model. This and previous studies 
have shown that calculus has a low capacity for retention of starch granules, 
an effect that is compounded by diagenetic effects in archaeological remains, 
resulting in low overall counts of extracted granules. 
The proportion of starches consumed will in many cases be reflected 
in the quantity
of starches extracted from the dental calculus---i.e., the more starch granules
entering the oral cavity, the more will be recovered from extraction---at 
least in modern calculus samples unaffected by diagenesis and hydrolysis.
Whether or not this also applies to archaeological samples remains to be tested.
Additionally, we have
shown that the size of granules will influence the likelihood of incorporation, 
as large (>20 $\mu$m) starches have a decreased incorporation rate, medium 
(10--20 $\mu$m)
starches an increased rate, and small (<10 $\mu$m) granules remained somewhat
constant. The size of calculus deposit also seems to influence the capacity of 
granule incorporation; as the size of the deposit increases, so does the
absolute count of incorporated granules.  
While we have shown multiple factors that influence the likelihood 
of incorporation, the process still appears to be somewhat stochastic. Further
research is needed to make sense of the contributing factors, and to explore the
mechanisms of intra-oral starch incorporation and retention in dental calculus.
The oral biofilm model described in this study provides a method
to explore the incorporation and extraction of dietary compounds from dental calculus 
in a controlled laboratory setting. We do not expect our model to replace *in vivo*
methods; instead, it can provide a complementary means to address the limitations
of *in vivo* studies, and unearth the potential biases associated 
with dietary research conducted on archaeological dental calculus.

### 2022-06-02 analysis/paper/paper.Rmd
incorporate reviewer comments and resubmit

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  #fig.path = "frontiers/figures/",
  fig.path = "../figures/",
  fig.align = "center",
  #fig.ext = "jpg",
  fig.pos = "H",
  dpi = 300) 
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(list.files(pattern = ".Rmd"))
#cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
# ## end refs.bib ##
# 
# ## Word count ##
# wd_count <- wordcountaddin::word_count()
cor_r <- signif(count_cor$estimate, 3)
cor_ci <- signif(count_cor$conf.int, 3)
cor2_r <- signif(count_cor$estimate, 3)
cor2_ci <- signif(count_cor$conf.int, 3)
```

@@

```{r microscope-fig, fig.cap="Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\\mu$m.", out.width="50%"}
knitr::include_graphics(here("analysis/figures/starches_w_bar.jpg"))
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  #fig.path = "frontiers/figures/",
  fig.path = "../figures/",
  fig.align = "center",
  #fig.ext = "jpg",
  fig.pos = "H",
  dpi = 300) 
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(list.files(pattern = ".Rmd"))
#cite_keys <- unlist(cite_keys, use.names = FALSE)
try(
  rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
  #rbbt::bbt_update_bib(here("analysis/paper/paper.Rmd"), here("analysis/paper/refs.bib"))
)
# ## end refs.bib ##
# 
# ## Word count ##
# wd_count <- wordcountaddin::word_count()
cor_r <- signif(count_cor$estimate, 3)
cor_ci <- signif(count_cor$conf.int, 3)
cor2_r <- signif(count_cor$estimate, 3)
cor2_ci <- signif(count_cor$conf.int, 3)
```

@@

```{r microscope-fig, fig.cap="Microscope images of biofilm samples that were exposed to the starch solutions. Starch granules can be seen within bacterial communities and isolated. Scale bar = 20 $\\mu$m.", fig.show='hold', out.width="40%"}
knitr::include_graphics(c(here("analysis/figures/starches_w_bar.jpg"), here("analysis/figures/st2C3.2-mix.jpg"), here("analysis/figures/st1B4-wheat.jpg"), here("analysis/figures/2D2-potato.jpg")))
```

### 2022-06-02 analysis/paper/paper.tex
incorporate reviewer comments and resubmit

Dental calculus has proven to contain a wealth of information on the
dietary habits of past populations. These insights have, to a large
extent, been obtained by the extraction and identification of starch
granules contained within the mineralised dental plaque from a wide
range of regions and time periods. The scope of previous studies have
been limited to microfossil extraction and identification to reconstruct
dietary preferences from the archaeological record, and few studies have
attempted to address the biases of starch retention in dental calculus.
Those that have considered this problem have been limited to \emph{in
vivo} studies on modern humans and non-human primates. Here, we present
a multispecies oral biofilm model, which allows experimental research on
starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were
exposed to treatment solutions with known quantities of dietary starches
(wheat and potato) during the 25-day growth period. After this, the
starch granules were extracted from the mature biofilm (by dissolution
in EDTA), and counted. We show that the granule counts extracted from
the model dental calculus represented a low proportion (ranging from
0.06\% to 0.16\%) of the total number of granules exposed to the
biofilms throughout the experiment. Additionally, we found that the
ratios of granule sizes from the extracted starch granules differed from
the original treatment solutions, with large granules (\textgreater20
\(\mu\)m) consistently being under-represented. We also found a
correlation between the absolute granule counts and dry-weight of the
biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
the concentration (count per mg) of granules and dry-weight (\emph{r} =
0.3, 90\%CI{[}0.0618, 0.506{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that
dental calculus presents a very small, and partly biased picture of the
original dietary intake of starches, with an over-representation of
plants producing granules smaller than 20 \(\mu\)m in size. The
experimental model presented here is well-suited to address the need for
further validation of methods and biases associated with dietary
research on dental calculus.

@@

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited (Radini et al., 2017), as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues (2017), specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans (Leonard et al., 2015)
and non-human primates (Power et al., 2015),
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved in this kind of \emph{in vivo} research. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals (Jepsen et al., 2011; Haffajee et al., 2009),
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals (Froehlich et al., 1987; Nater et al., 2005),
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being
explored (García-Granero, 2020), and it becomes clear that using
dental calculus to reconstruct diet is a highly unpredictable process.

@@

Statistical analysis was conducted in R version 4.1.3 (2022-03-10) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
maximum length were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(van de Velde et al., 2002; Jane et al., 1994; Reichert, 1913, 364--365; Seidemann, 1966, 174--176).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between biofilm size and retained
starch granules (Figure \ref{fig:cor-plot}A);
a result that contradicts findings from archaeological contexts
(Wesolowski et al., 2010; Dudgeon and Tromp, 2014).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot}B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = 5.53 mg) were on average larger than mixed treatment samples (mean = 4.28 mg) (Table \ref{tab:anova-tbl}); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

While we are unable to sufficiently address the
mechanism(s) of starch incorporation with the data obtained in this study,
the dental calculus model presented here is uniquely suited to explore
these questions and may improve interpretations of dietary practices in past
populations. Further analyses using this model can address the call for more
baseline testing of biases associated with dietary research conducted on dental calculus
(Le Moyne and Crowther, 2021).
Our high-throughput experimental setup allows us a
higher degree of control over the factors that influence starch incorporation and
retention, such as dietary intake, differential survivability of starches,
and inter- and intra-individual variation in plaque accumulation and mineralisation.
The latter is especially difficult to control \emph{in vivo} as it is influenced by
numerous factors including genetics, diet, salivary flow, and tooth position and
morphology
(Jepsen et al., 2011; Simón-Soro et al., 2013; Proctor et al., 2018; Fagernäs et al., 2021; Haffajee et al., 2009),
as well as evolutionary differences (Fellows Yates et al., 2021).
It can also facilitate training of students and researchers on methods of
dental calculus analysis, such as starch and phytolith extraction and
identification, where it can replace the use of finite archaeological resources.

@@

This preliminary study shows that a very small proportion of the input starch
granules are retained in a dental calculus model. This and previous studies
have shown that calculus has a low capacity for retention of starch granules,
an effect that is compounded by diagenetic effects in archaeological remains,
resulting in low overall counts of extracted granules.
The proportion of starches consumed will in many cases be reflected
in the quantity
of starches extracted from the dental calculus---i.e., the more starch granules
entering the oral cavity, the more will be recovered from extraction---at
least in modern calculus samples unaffected by diagenesis and hydrolysis.
Whether or not this also applies to archaeological samples remains to be tested.
Additionally, we have
shown that the size of granules will influence the likelihood of incorporation,
as large (\textgreater20 \(\mu\)m) starches have a decreased incorporation rate, medium
(10--20 \(\mu\)m)
starches an increased rate, and small (\textless10 \(\mu\)m) granules remained somewhat
constant. The size of calculus deposit also seems to influence the capacity of
granule incorporation; as the size of the deposit increases, so does the
absolute count of incorporated granules.\\
While we have shown multiple factors that influence the likelihood
of incorporation, the process still appears to be somewhat stochastic. Further
research is needed to make sense of the contributing factors, and to explore the
mechanisms of intra-oral starch incorporation and retention in dental calculus.
The oral biofilm model described in this study provides a method
to explore the incorporation and extraction of dietary compounds from dental calculus
in a controlled laboratory setting, and unearth the potential biases associated
with dietary research conducted on archaeological dental calculus.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-R-rrtools}{}}%
Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research compendium}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-R-here}{}}%
Müller, K. (2020). \emph{Here: {A} simpler way to find your files}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-R-patchwork}{}}%
Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-R-broom}{}}%
Robinson, D., Hayes, A., and Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-sissonsMultistationDentalPlaque1991}{}}%
Sissons, C. H., Cutress, T. W., Hoffman, M. P., and Wakefield, J. S. J. (1991). A {Multi-station Dental Plaque Microcosm} ({Artificial Mouth}) for the {Study} of {Plaque Growth}, {Metabolism}, {pH}, and {Mineralization}: \emph{Journal of Dental Research}. doi:\href{https://doi.org/10.1177/00220345910700110301}{10.1177/00220345910700110301}.

@@

{\centering \includegraphics[width=0.7\linewidth]{/mnt/hogwarts/Insync/publications/PhD/byocstarch/analysis/figures/protocol_overview}

@@

{\centering \includegraphics[width=0.5\linewidth]{/mnt/hogwarts/Insync/publications/PhD/byocstarch/analysis/figures/starches_w_bar}

@@

\caption{Microscope image of a biofilm sample exposed to the potato starch solution. Potato granules can be seen within a bacterial community. Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
\end{figure}

>>>>

Dental calculus has proven to contain a wealth of information on the
dietary habits of past populations. These insights have, to a large
extent, been obtained by the extraction and identification of starch
granules contained within the mineralised dental plaque from a wide
range of regions and time periods. The scope of previous studies have
been limited to microfossil extraction and identification to reconstruct
dietary preferences from the archaeological record, and few studies have
attempted to address the biases of starch retention in dental calculus.
Those that have considered this problem have been limited to \emph{in
vivo} studies on modern humans and non-human primates. Here, we present
a multispecies oral biofilm model, which allows experimental research on
starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were
exposed to treatment solutions with known quantities of dietary starches
(wheat and potato) during the 25-day growth period. After this, the
starch granules were extracted from the mature biofilm (by dissolution
in EDTA), and counted. We show that the granule counts extracted from
the model dental calculus represented a low proportion (ranging from
0.06\% to 0.16\%) of the total number of granules exposed to the
biofilms throughout the experiment. Additionally, we found that the
ratios of granule sizes from the extracted starch granules differed from
the original treatment solutions, with large granules (\textgreater20
\(\mu\)m) consistently being under-represented. We also found a
positive correlation between the absolute granule counts and dry-weight of the
biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), meaning the absolute
quantity of starch granules will increase as the size of the calculus deposit
increases. A similar, but weaker correlation was found between
the concentration (count per mg) of granules and dry-weight (\emph{r} =
0.3, 90\%CI{[}0.0618, 0.506{]}).\\
Our results complement and reinforce previous \emph{in vivo} studies suggesting
that dental calculus presents a very small, and partly biased picture of the
original dietary intake of starches, with an over-representation of
plants producing granules smaller than 20 \(\mu\)m in size. The
experimental model presented here is well-suited to address the need for
further validation of methods and biases associated with dietary
research on dental calculus.

@@

Despite this, our knowledge of dental calculus and the incorporation pathways of
the various markers is limited (Radini et al., 2017), as is our
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested
by Tromp and colleagues (2017), specifically regarding
decalcification using HCl or EDTA.
The authors found significantly more starches with the EDTA extraction method
than the HCl extraction method; however, as noted by the authors, comparisons
involving archaeological calculus are problematic due to variability between and
within individuals.
Studies conducted on modern humans (Leonard et al., 2015)
and non-human primates (Power et al., 2015; Power et al., 2021)
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited,
despite meticulous documentation and observation, due to unknown variables and
uncertainty involved in this kind of \emph{in vivo} research. Dental calculus is a complex
oral biofilm with a multifactorial aetiology and variable formation rates both
within and between individuals (Jepsen et al., 2011; Haffajee et al., 2009),
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral \(\alpha\)-amylase differs both between and
within individuals (Froehlich et al., 1987; Nater et al., 2005),
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing (Hardy et al., 2018), as well as post-depositional
processes that are still being explored
(García-Granero, 2020; Mercader et al., 2018),
and it becomes clear that using
dental calculus to reconstruct diet is a highly unpredictable process.

@@

Statistical analysis was conducted in R version 4.2.0 (2022-04-22) (R Core Team, 2020) and
the following packages: tidyverse (Wickham et al., 2019), broom (Robinson et al., 2021),
here (Müller, 2020), and patchwork (Pedersen, 2020).

@@

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 \(\mu\)m in
maximum length were underrepresented in the calculus samples compared to the original
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 \(\mu\)m) were often over-represented
(Table \ref{tab:sample-prop-tbl},
and
Figure \ref{fig:ratio-plots}).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 \(\mu\)m in maximum length, whereas wheat
granules generally only reach up to 35 \(\mu\)m
(Gismondi et al., 2019; Haslam, 2004; Seidemann, 1966, 174--176).
Granule morphology may also play a role. Large wheat granules
are lenticular and have a larger surface area compared
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area
(van de Velde et al., 2002; Jane et al., 1994; Reichert, 1913, 364--365; Seidemann, 1966, 174--176).
Another potentially important factor
from our results is the size of the calculus deposit. We found a
strong positive correlation between size of biofilm deposit and
retained starch granules (Figure \ref{fig:cor-plot}A), meaning larger calculus
deposits contain a higher quantity of granules;
a result that contradicts findings from archaeological contexts
(Wesolowski et al., 2010; Dudgeon and Tromp, 2014).
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \ref{fig:cor-plot}B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = 5.53 mg) were on average larger than mixed treatment samples (mean = 4.28 mg) (Table \ref{tab:anova-tbl}); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

@@

While we are unable to sufficiently address the
mechanism(s) of starch incorporation with the data obtained in this study,
the dental calculus model presented here is uniquely suited to explore
these questions and may improve interpretations of dietary practices in past
populations. Further analyses using this model can address the call for more
baseline testing of biases associated with dietary research conducted on dental calculus
(Le Moyne and Crowther, 2021).
Our high-throughput experimental setup allows us a
higher degree of control over the factors that influence starch incorporation and
retention, such as dietary intake, differential survivability of starches,
and inter- and intra-individual variation in plaque accumulation and mineralisation.
The latter is especially difficult to control \emph{in vivo} as it is influenced by
numerous factors including genetics, diet, salivary flow, and tooth position and
morphology
(Jepsen et al., 2011; Simón-Soro et al., 2013; Proctor et al., 2018; Fagernäs et al., 2021; Haffajee et al., 2009),
as well as evolutionary differences (Fellows Yates et al., 2021). The set of
limitations for our model differ from \emph{in vivo} methods and, as such, we expect
our model to complement the results and interpretations of existing and new
\emph{in vivo} studies.
It can also facilitate training of students and researchers on methods of
dental calculus analysis, such as starch and phytolith extraction and
identification, where it can replace the use of finite archaeological resources.

@@

This preliminary study shows that a very small proportion of the input starch
granules are retained in a dental calculus model. This and previous studies
have shown that calculus has a low capacity for retention of starch granules,
an effect that is compounded by diagenetic effects in archaeological remains,
resulting in low overall counts of extracted granules.
The proportion of starches consumed will in many cases be reflected
in the quantity
of starches extracted from the dental calculus---i.e., the more starch granules
entering the oral cavity, the more will be recovered from extraction---at
least in modern calculus samples unaffected by diagenesis and hydrolysis.
Whether or not this also applies to archaeological samples remains to be tested.
Additionally, we have
shown that the size of granules will influence the likelihood of incorporation,
as large (\textgreater20 \(\mu\)m) starches have a decreased incorporation rate, medium
(10--20 \(\mu\)m)
starches an increased rate, and small (\textless10 \(\mu\)m) granules remained somewhat
constant. The size of calculus deposit also seems to influence the capacity of
granule incorporation; as the size of the deposit increases, so does the
absolute count of incorporated granules.\\
While we have shown multiple factors that influence the likelihood
of incorporation, the process still appears to be somewhat stochastic. Further
research is needed to make sense of the contributing factors, and to explore the
mechanisms of intra-oral starch incorporation and retention in dental calculus.
The oral biofilm model described in this study provides a method
to explore the incorporation and extraction of dietary compounds from dental calculus
in a controlled laboratory setting. We do not expect our model to replace \emph{in vivo}
methods; instead, it can provide a complementary means to address the limitations
of \emph{in vivo} studies, and unearth the potential biases associated
with dietary research conducted on archaeological dental calculus.

@@

\leavevmode\vadjust pre{\hypertarget{ref-hardyRecoveringInformation2018}{}}%
Hardy, K., Buckley, S., and Copeland, L. (2018). Pleistocene dental calculus: {Recovering} information on {Paleolithic} food items, medicines, paleoenvironment and microbes. \emph{Evolutionary Anthropology: Issues, News, and Reviews} 27, 234--246. doi:\href{https://doi.org/10.1002/evan.21718}{10.1002/evan.21718}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Rrrtools}{}}%
Marwick, B. (2019). \emph{Rrtools: {Creates} a reproducible research compendium}.

\leavevmode\vadjust pre{\hypertarget{ref-mercaderExaggeratedExpectations2018}{}}%
Mercader, J., Akeju, T., Brown, M., Bundala, M., Collins, M. J., Copeland, L., Crowther, A., Dunfield, P., Henry, A., Inwood, J., et al. (2018). Exaggerated expectations in ancient starch research and the need for new taphonomic and authenticity criteria. \emph{FACETS} 3, 777--798. doi:\href{https://doi.org/10.1139/facets-2017-0126}{10.1139/facets-2017-0126}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Rhere}{}}%
Müller, K. (2020). \emph{Here: {A} simpler way to find your files}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Rpatchwork}{}}%
Pedersen, T. L. (2020). \emph{Patchwork: {The} composer of plots}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-powerRepresentativenessDental2021}{}}%
Power, R. C., Wittig, R. M., Stone, J. R., Kupczik, K., and Schulz-Kornas, E. (2021). The representativeness of the dental calculus dietary record: Insights from {Taï} chimpanzee faecal phytoliths. \emph{Archaeological and Anthropological Sciences} 13, 104. doi:\href{https://doi.org/10.1007/s12520-021-01342-z}{10.1007/s12520-021-01342-z}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-Rbroom}{}}%
Robinson, D., Hayes, A., and Couch, S. (2021). \emph{Broom: {Convert} statistical objects into tidy tibbles}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-sissonsMultistationPlaque1991}{}}%
Sissons, C. H., Cutress, T. W., Hoffman, M. P., and Wakefield, J. S. J. (1991). A {Multi-station Dental Plaque Microcosm} ({Artificial Mouth}) for the {Study} of {Plaque Growth}, {Metabolism}, {pH}, and {Mineralization}: \emph{Journal of Dental Research}. doi:\href{https://doi.org/10.1177/00220345910700110301}{10.1177/00220345910700110301}.

@@

{\centering \includegraphics[width=0.7\linewidth]{../figures/protocol_overview}

@@

{\centering \includegraphics[width=0.4\linewidth]{../figures/starches_w_bar} \includegraphics[width=0.4\linewidth]{../figures/st2C3.2-mix} \includegraphics[width=0.4\linewidth]{../figures/st1B4-wheat} \includegraphics[width=0.4\linewidth]{../figures/2D2-potato}

@@

\caption{Microscope images of biofilm samples that were exposed to the starch solutions. Starch granules can be seen within bacterial communities and isolated. Scale bar = 20 $\mu$m.}\label{fig:microscope-fig}
\end{figure}

### 2022-06-02 analysis/supplementary-materials/supp-mat.Rmd
incorporate reviewer comments and resubmit

```{r img-2, fig.cap="Microscope image of wheat starch from a Potato treatment sample.", echo=FALSE, out.width="80%"}
knitr::include_graphics(here("analysis/supplementary-materials/figures/SNAP-103412-0006.jpg"))
```

>>>>

```{r img-2, fig.cap="Microscope image of potato starch from a potato treatment sample.", echo=FALSE, out.width="80%"}
knitr::include_graphics(here("analysis/supplementary-materials/figures/SNAP-103412-0006.jpg"))
```
